---
title: "results_so_far"
author: "Jon Minton"
date: "24/02/2022"
output: ioslides_presentation
---





```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Data 

- England & Wales (`GBRTENW`)
- Scotland (`GBR_SCO`)
- UK as a whole (`GBR_UK`)
- France (`FRANTP`)
- Spain (`ESP`)
- Italy (`ITA`)
- Germany
   - Total Germany (`DEUTNP`)
   - East Germany (`DEUTE`)
   - West Germany (`DEUTW`)
   - Simulated/Synthetic Germany
 

```{r, echo=FALSE}
# load packages

pacman::p_load(here, tidyverse)

# load data 

hmd_lt <- read_rds(here("data", "lifetables.rds"))

```

## Analysis 

For each of these countries, and for males and females separately, we are interested in the annual changes in life expectancy at birth ($e_0$) and life expectancy at age 65 ($e_{65}$), from 1980 to the last available year for each country. 

```{r, echo = FALSE}

countries_of_interest <- c(
  "GBRTENW",
  "GBR_SCO",
  "GBR_UK",
  "FRATNP",
  "ESP",
  "ITA",
  "DEUTNP",
  "DEUTE", 
  "DEUTW"
)

hmd_ex_selected_countries <- 
  hmd_lt %>% 
    filter(code %in% countries_of_interest) %>% 
    filter(Age %in% c(0, 65)) %>% 
    select(code, Year, Age, sex, ex) %>% 
    rename(x = Age, year = Year)

hmd_ex_selected_countries




```

## Life expectancy at birth


```{r, echo = FALSE, fig.cap = "Life Expectancy at birth"}

hmd_ex_selected_countries %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~code) +
  labs(
    x = "Year",
    y = "Life Expectancy at birth",
    title = "Life expectancies at birth for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )



```


## life expectancy at age 65

```{r, echo = FALSE, fig.cap = "Life Expectancy at age 65"}
hmd_ex_selected_countries %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~code) +
  labs(
    x = "Year",
    y = "Life Expectancy at age 65",
    title = "Life expectancies at age 65 for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )

```




