---
title: "Life Expectancies for selected populations"
author: "Jon Minton"
date: '2022-07-23'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Aim

This appendix will show the life expectancies for the populations compared in the paper. 


# Data 

```{r}
library(tidyverse)

# load data 
hmd_lt <- read_rds("https://github.com/JonMinton/change-in-ex/blob/main/data/lifetables.rds?raw=true")


# Labels for codes 
country_code_lookup <- 
  tribble(
    ~code, ~country,
    "DEUTNP", "Germany",
    "DEUTE", "East Germany",
    "DEUTW", "West Germany",
    "ESP", "Spain",
    "FRATNP", "France", 
    "ITA", "Italy",
    "GBRTENW", "England & Wales",
    "GBR_SCO", "Scotland",
    "DEUTSYNTH", "Synthetic Germany",
    "NLD", "Netherlands"
  )

countries_of_interest <- c(
  "GBRTENW",
  "GBR_SCO",
  "GBR_UK",
  "FRATNP",
  "ESP",
  "ITA",
  "DEUTNP",
  "DEUTE", 
  "DEUTW",
  "NLD"
)

source("https://raw.githubusercontent.com/JonMinton/change-in-ex/main/R/make_synthetic_germany_functions.R")

source("https://raw.githubusercontent.com/JonMinton/change-in-ex/main/R/make_pop_selection.R")

```

# Graphs 

## Life Expectancy at birth

```{r}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~country) +
  labs(
    x = "Year",
    y = "Life Expectancy at birth",
    title = "Life expectancies at birth for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )
```

Equivalently


```{r}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  ggplot(aes(x = year, y = ex, group = country, colour = country, linetype = country)) + 
  geom_line() +
  facet_wrap(~sex) +
  labs(
    x = "Year",
    y = "Life Expectancy at birth",
    title = "Life expectancies at birth for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )

```

## Life expectancy at age 65 


```{r}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~country) +
  labs(
    x = "Year",
    y = "Life Expectancy at age 65",
    title = "Life expectancies at age 65 years  for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )
```

Equivalently: 

```{r}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year, y = ex, group = country, colour = country, linetype = country)) + 
  geom_line() +
  facet_wrap(~sex) +
  labs(
    x = "Year",
    y = "Life Expectancy at age 65",
    title = "Life expectancies at age 65 years  for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )
```
# Average life expectancy change 

The following shows the average change in life expectancy for each population over the full period, and including East Germany and West Germany as separate populations


```{r}
hmd_ex_selected_countries_with_synth %>% 
    filter(year >= 1979) %>% 
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    nest() %>% 
    mutate(
      rwd_model = map(data, ~lm(delta_ex ~ 1, data = . )),
      ar_model  = map(data, ~lm(delta_ex ~ lag(delta_ex), data = .))
    ) %>% 
    mutate(
      aic_rwd = map_dbl(rwd_model, AIC),
      aic_ar  = map_dbl(ar_model, AIC)
    ) %>% 
    mutate(
      diff_aic = aic_ar - aic_rwd
    )  %>% 
  # filter(diff_aic > 0)
  mutate(
    tidied_ar_model = map(ar_model, broom::tidy)
  ) %>% 
  unnest(tidied_ar_model) %>% 
  filter(term == "(Intercept)") %>% 
  arrange(desc(estimate)) %>%
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  select(code, country, x, sex, estimate, std.error) %>% 
  mutate(display_label = glue::glue("{country}_{x}_{sex}")) %>% 
  mutate(estimate_weeks = estimate * 52.25) %>% 
  ggplot(aes(estimate_weeks, fct_reorder(display_label, estimate_weeks))) + 
  geom_point(aes(colour = x), show.legend = FALSE) +
  expand_limits( x = 0) + 
  geom_vline(xintercept = 0) + 
  labs(
    x = "Average annual life expectancy change",
    y = "Country-sex combination",
    title = "Average annual improvements by country and sex",
    subtitle = "Range: 1980 to 2020 or nearest available years",
    caption = "Source: Human Mortality Database"
  ) + 
  facet_wrap(~x)

```
The figure below shows the same for those populations considered in the main paper.

```{r}
hmd_ex_selected_countries_with_synth %>% 
    filter(year >= 1979) %>% 
    filter(!(code %in% c("DEUTE", "DEUTW", "DEUTNP"))) %>%  # Using only synthetic germany for longer time period
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    nest() %>% 
    mutate(
      rwd_model = map(data, ~lm(delta_ex ~ 1, data = . )),
      ar_model  = map(data, ~lm(delta_ex ~ lag(delta_ex), data = .))
    ) %>% 
    mutate(
      aic_rwd = map_dbl(rwd_model, AIC),
      aic_ar  = map_dbl(ar_model, AIC)
    ) %>% 
    mutate(
      diff_aic = aic_ar - aic_rwd
    )  %>% 
  # filter(diff_aic > 0)
  mutate(
    tidied_ar_model = map(ar_model, broom::tidy)
  ) %>% 
  unnest(tidied_ar_model) %>% 
  filter(term == "(Intercept)") %>% 
  arrange(desc(estimate)) %>%
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>%
  filter(!is.na(country)) %>% 
  select(code, country, x, sex, estimate, std.error) %>% 
  mutate(display_label = glue::glue("{country}_{x}_{sex}")) %>% 
  mutate(estimate_weeks = estimate * 52.25) %>% 
  ggplot(aes(estimate_weeks, fct_reorder(display_label, estimate_weeks))) + 
  geom_point(aes(colour = x), show.legend = FALSE) +
  expand_limits( x = 0) + 
  geom_vline(xintercept = 0) + 
  labs(
    x = "Average annual life expectancy change (weeks per year)",
    y = "Country-sex combination",
    title = "Average annual improvements by country and sex",
    subtitle = "Range: 1980 to 2020 or nearest available years",
    caption = "Source: Human Mortality Database"
  ) + 
  facet_wrap(~x)


```

